---
title: "Mutation frequency analysis"
author: Barbara Arbeithuber and Marzia A. Cremona
output: 
  html_notebook: 
    toc: yes
    toc_float: true
    toc_depth: 3
---

# Data preparation and correlation plot

```{r message=FALSE, warning=FALSE}
library(ggplot2)
library(reshape2)
library(plyr)
library(readxl)
require(lme4)
require(MuMIn)
require(multcomp)
```

Remove sites:
255 G>C (because of insertion - microsatellite)
273 C>A (end of read)
11022 A>C (beginning of polyA - coupled with deletion of a A)
297 A>C (beginning of polyC - coupled with insertion of a C)
13750 A>C (in polyC stretch)
11029 A>C (in polyC stretch)

```{r}
mut2 <- read_excel("2020-09_Rh_mut_freqs.xlsx", sheet = "Rh_mut_freq")
head(mut2, n=2)
nrow(mut2)
```

```{r}
# select only samples with depth >= 100x
mut <- subset(mut2, mean_depth_DCS >= 100)
nrow(mut)
```
```{r fig.height=6, fig.width=11}
options(repr.plot.width=11, repr.plot.height=6)

p <- ggplot(mut, aes(x=age_years, y=(n_mutations/nt_sequenced), color=tissue)) + 
  geom_point(shape=16, size=4, alpha=0.9) +
  geom_smooth(method=lm, formula= y~x, se = FALSE) +
  scale_color_manual(values=c("red","blue", "green", "black")) +
# scale_y_continuous(limits = c(0, 0.0000034),breaks=c(seq(0,0.0000034,0.0000005))) +
  theme_classic() +
  labs(title="",x="Age [years]",y="Mutation frequency [substitutions/bp]") +
# scale_fill_manual(values=c("red","navy","grey","grey20"), labels=c("Brain", "Skeletal muscle", "Oocyte", "Oocyte pool")) +
# scale_color_manual(values=c("red","navy","grey","grey20"), labels=c("Brain", "Skeletal muscle", "Oocyte", "Oocyte pool")) +
  theme(axis.text.x = element_text(size=20, color="black"),axis.text.y = element_text(size=20, color="black"), axis.title.x = element_text(size=20), axis.title.y = element_text(size=20)) +
  guides(color=guide_legend(title="Tissue type")) +
  theme(legend.text=element_text(size=20)) +
  theme(legend.title=element_text(size=20, face="bold")) +
  theme(plot.title = element_text(size=20, lineheight=.8, face="bold", hjust = 0.5)) 

ggsave(filename="2020-10-14_mut_freq_d100.pdf", plot=p, , width = 11, height = 6)

p
```





# Mixed-effects linear model

Transform in factor (set muscle as baseline tissue), rescale age_years
```{r}
mut$tissue = factor(mut$tissue, levels = c('muscle', 'heart', 'liver', 'oocyte'))
mut$macaque_ID = as.factor(mut$macaque_ID)
mut$age_years_scaled = as.numeric(scale(mut$age_years))
```

Model Age vs Mutation frequency taking into consideration tissue type (fixed effect) and individual (random effect). 
Note: in a binomial model prior weights (`weights = nt_sequenced`) are used to give the number of trials when the response is the proportion of successes (`mut_freq`):
```{r}
glmer_results <- glmer(mut_freq ~ age_years_scaled*tissue + (1|macaque_ID), 
                       weights = nt_sequenced, family = binomial(link = "logit"), 
                       control=glmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)), 
                       data = mut)
summary(glmer_results)
```

Coefficients transforming back `age_years` to the original scale:
```{r}
beta_model = summary(glmer_results)$coefficients[,1]
mu = mean(mut$age_years)
sigma = sd(mut$age_years)
beta_rescaled_back = beta_model
beta_rescaled_back[c(2,6:8)] = beta_model[c(2,6:8)]/sigma
beta_rescaled_back[c(1,3:5)] = beta_model[c(1,3:5)] - beta_model[c(2,6:8)]*mu/sigma

as.matrix(beta_rescaled_back)
```

Odds ratio:
```{r}
odds_ratio = exp(beta_rescaled_back)
odds_ratio[1] = NA
as.matrix(odds_ratio)
```


Check residuals
```{r}
plot(glmer_results, col=c("green", "red", "blue", "black")[mut$tissue])
```

Compute pseudo_R2
```{r}
glmer_null <- glmer(mut_freq ~ (1|macaque_ID), 
                    weights = nt_sequenced, family = binomial(link = "logit"), 
                    control=glmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)), 
                    data = mut)

r.squaredGLMM(glmer_results, null = glmer_null)
r2_full <- r.squaredGLMM(glmer_results, null = glmer_null)[1,1]
r2_full
```

## Test random effect
Test random effect macaque_ID
```{r}
glmer_red <- glm(mut_freq ~ age_years_scaled*tissue, 
                 weights = nt_sequenced, family = binomial(link = "logit"), 
                 data = mut)
anova(glmer_results, glmer_red)
anova(glmer_results, glmer_red)$`Pr(>Chisq)`
```

Conditional modes of the random effects:
```{r fig.height=5, fig.width=6}
ranef_macaque_ID = ranef(glmer_results)$macaque_ID[,1]
summary(ranef_macaque_ID)

layout(matrix(c(1,1,2), nrow = 3))
hist(ranef_macaque_ID, main = 'Random effect macaque_ID', xlab = 'Conditional mode', xlim = c(-0.4, 0.4))
boxplot(ranef_macaque_ID, main = 'Random effect macaque_ID', xlab = 'Conditional mode', horizontal = TRUE, ylim = c(-0.4, 0.4))
```

Test if a random effect primate_center would also be important
```{r}
glmer_center <- glmer(mut_freq ~ age_years_scaled*tissue + (1|macaque_ID) + (1|primate_center), 
                      weights = nt_sequenced, family = binomial(link = "logit"), 
                      control=glmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)), 
                      data = mut)
anova(glmer_center, glmer_results)
```
It's not important at all (once macaque_ID is included)


## Test fixed effects
Test effect of age_years
```{r}
glmer_red <- glmer(mut_freq ~ tissue + (1|macaque_ID), 
                   weights = nt_sequenced, family = binomial(link = "logit"), 
                   control=glmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)), 
                   data = mut)
a=anova(glmer_results, glmer_red)
a$`Pr(>Chisq)`
```
Partial R2
```{r}
r2 <- r.squaredGLMM(glmer_red, null=glmer_null)[1,1]
(r2_full - r2) / (1-r2) * 100
```

Test effect of tissue
```{r}
glmer_red <- glmer(mut_freq ~ age_years_scaled + (1|macaque_ID), 
                   weights = nt_sequenced, family = binomial(link = "logit"), 
                   control=glmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)), 
                   data = mut)
a=anova(glmer_results, glmer_red)
a$`Pr(>Chisq)`
```
Partial R2
```{r}
r2 <- r.squaredGLMM(glmer_red, null=glmer_null)[1,1]
(r2_full - r2) / (1-r2) * 100
```

Test effect of different slopes for different tissues
```{r}
glmer_red <- glmer(mut_freq ~ age_years_scaled + tissue + (1|macaque_ID), 
                   weights = nt_sequenced, family = binomial(link = "logit"), 
                   control=glmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)), 
                   data = mut)
a=anova(glmer_results, glmer_red)
a$`Pr(>Chisq)`
```

Test effect of different intercepts for different tissues
```{r}
glmer_red <- glmer(mut_freq ~ age_years_scaled + age_years_scaled:tissue + (1|macaque_ID), 
                   weights = nt_sequenced, family = binomial(link = "logit"), 
                   control=glmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)), 
                   data = mut)
a=anova(glmer_results, glmer_red)
a$`Pr(>Chisq)`
```

## Test slopes for different tissues
```{r}
coeffs_glmer = coeffs(glmer_results)
coeffs_glmer
```

Test muscle slope:
```{r}
beta1 = coeffs_glmer[2]
beta1/sigma

K = matrix(0, nrow = 1, ncol = length(coeffs_glmer))
K[1,c(2)] = 1/sigma
colnames(K) = names(coeffs_glmer)
rownames(K) = "slope_muscle"
summary(glht(glmer_results, linfct = K))
2*pnorm(-summary(glht(glmer_results, linfct = K))$test$tstat)
```

Test heart slope:
```{r}
beta1_5 = coeffs_glmer[2]+coeffs_glmer[6]
beta1_5/sigma

K = matrix(0, nrow = 1, ncol = length(coeffs_glmer))
K[1,c(2,6)] = 1/sigma
colnames(K) = names(coeffs_glmer)
rownames(K) = "slope_heart"
summary(glht(glmer_results, linfct = K))
2*pnorm(-summary(glht(glmer_results, linfct = K))$test$tstat)
```

Test liver slope:
```{r}
beta1_6 = coeffs_glmer[2]+coeffs_glmer[7]
beta1_6/sigma

K = matrix(0, nrow = 1, ncol = length(coeffs_glmer))
K[1,c(2,7)] = 1/sigma
colnames(K) = names(coeffs_glmer)
rownames(K) = "slope_liver"
summary(glht(glmer_results, linfct = K))
2*pnorm(-summary(glht(glmer_results, linfct = K))$test$tstat)
```

Test oocyte slope:
```{r}
beta1_7 = coeffs_glmer[2]+coeffs_glmer[8]
beta1_7/sigma

K = matrix(0, nrow = 1, ncol = length(coeffs_glmer))
K[1,c(2,8)] = 1/sigma
colnames(K) = names(coeffs_glmer)
rownames(K) = "slope_oocyte"
summary(glht(glmer_results, linfct = K))
2*pnorm(-summary(glht(glmer_results, linfct = K))$test$tstat)
```


## Plot model prediction

```{r fig.height=7, fig.width=11}
options(repr.plot.width=11, repr.plot.height=7)

age_plot = seq(min(mut$age_years),max(mut$age_years),length.out = 50)
data_plot = data.frame(tissue = rep(levels(mut$tissue), each=50),
                       age_years = rep(age_plot, 4),
                       age_years_scaled = rep((age_plot-mean(mut$age_years))/sd(mut$age_years), 4))
data_plot$mut_freq = predict(glmer_results, data_plot, re.form=~0, type='response')

p <- ggplot(mut, aes(x=age_years, y=mut_freq, color=tissue)) + 
  geom_point(shape=16, size=4, alpha=0.9) +
  scale_color_manual(values=c("green", "red","blue", "black")) +
  theme_classic() +
  labs(title="",x="Age [years]",y="Mutation frequency [substitutions/bp]") +
  theme(axis.text.x = element_text(size=20, color="black"),axis.text.y = element_text(size=20, color="black"), axis.title.x = element_text(size=20), axis.title.y = element_text(size=20)) +
  guides(color=guide_legend(title="Tissue type")) +
  theme(legend.text=element_text(size=20)) +
  theme(legend.title=element_text(size=20, face="bold")) +
  theme(plot.title = element_text(size=20, lineheight=.8, face="bold", hjust = 0.5)) +
  
  geom_point(mut, mapping = aes(x = age_years, y = fitted(glmer_results), color=tissue), shape = 3, size = 4) + 
  geom_line(data_plot, mapping = aes(x = age_years, y = mut_freq, color=tissue), lwd = 1.5)

ggsave(filename="2021-03-08_mut_freq_d100_model_pred.pdf", plot=p, , width = 11, height = 7)

p
```

```{r}
age_plot = seq(min(mut$age_years),max(mut$age_years),length.out = 50)
data_plot = data.frame(tissue = rep(levels(mut$tissue), each=50),
                       age_years = rep(age_plot, 4),
                       age_years_scaled = rep((age_plot-mean(mut$age_years))/sd(mut$age_years), 4))
data_plot$mut_freq = predict(glmer_results, data_plot, re.form=~0, type='response')

# bootstrap for computing CI using quantile approach
predFun <- function(fit){
  predict(fit, data_plot, re.form=~0, type='response')
}
merBoot = bootMer(glmer_results, FUN=predFun, nsim=100)
CIlevel = 0.95
data_plot$mut_freq_CIlower = apply(merBoot$t, 2, function(x) as.numeric(quantile(x, probs=(1-CIlevel)/2, na.rm=TRUE)))
data_plot$mut_freq_CIupper = apply(merBoot$t, 2, function(x) as.numeric(quantile(x, probs=(1+CIlevel)/2, na.rm=TRUE)))
```

```{r fig.height=7, fig.width=11}
options(repr.plot.width=11, repr.plot.height=7)

p <- ggplot(mut, aes(x=age_years, y=mut_freq, color=tissue)) + 
  geom_point(shape=16, size=4, alpha=0.9) +
  scale_color_manual(values=c("green", "red","blue", "black")) +
  theme_classic() +
  labs(title="",x="Age [years]",y="Mutation frequency [substitutions/bp]") +
  theme(axis.text.x = element_text(size=20, color="black"),axis.text.y = element_text(size=20, color="black"), axis.title.x = element_text(size=20), axis.title.y = element_text(size=20)) +
  guides(color=guide_legend(title="Tissue type")) +
  theme(legend.text=element_text(size=20)) +
  theme(legend.title=element_text(size=20, face="bold")) +
  theme(plot.title = element_text(size=20, lineheight=.8, face="bold", hjust = 0.5)) +
  
  geom_point(mut, mapping = aes(x = age_years, y = fitted(glmer_results), color=tissue), shape = 3, size = 4) + 
  geom_line(data_plot, mapping = aes(x = age_years, y = mut_freq, color=tissue), lwd = 1.5) +
  
  geom_ribbon(data_plot, mapping = aes(x=age_years, ymax=mut_freq_CIlower, ymin=mut_freq_CIupper), fill="gray", alpha=.25)

ggsave(filename="2021-12-23_mut_freq_d100_model_pred_CI.pdf", plot=p, , width = 11, height = 7)

p
```

```{r fig.height=7, fig.width=11}
options(repr.plot.width=11, repr.plot.height=7)

age_plot = seq(min(mut$age_years),max(mut$age_years),length.out = 50)
data_plot = data.frame(tissue = rep(levels(mut$tissue), each=50),
                       age_years = rep(age_plot, 4),
                       age_years_scaled = rep((age_plot-mean(mut$age_years))/sd(mut$age_years), 4))
data_plot$mut_freq = predict(glmer_results, data_plot, re.form=~0, type='response')

p <- ggplot(mut, aes(x=age_years, y=mut_freq, color=tissue)) + 
  geom_point(shape=16, size=4, alpha=0.9) +
  scale_color_manual(values=c("green", "red","blue", "black")) +
  theme_classic() +
  labs(title="",x="Age [years]",y="Mutation frequency [substitutions/bp]") +
  theme(axis.text.x = element_text(size=20, color="black"),axis.text.y = element_text(size=20, color="black"), axis.title.x = element_text(size=20), axis.title.y = element_text(size=20)) +
  guides(color=guide_legend(title="Tissue type")) +
  theme(legend.text=element_text(size=20)) +
  theme(legend.title=element_text(size=20, face="bold")) +
  theme(plot.title = element_text(size=20, lineheight=.8, face="bold", hjust = 0.5)) +
  
  #geom_point(mut, mapping = aes(x = age_years, y = fitted(glmer_results), color=tissue), shape = 3, size = 4) + 
  geom_line(data_plot, mapping = aes(x = age_years, y = mut_freq, color=tissue), lwd = 1.5)

ggsave(filename="2021-03-08_mut_freq_d100_model_pred_no_plus.pdf", plot=p, , width = 11, height = 7)

p
```



# Piecewise model only for oocytes, break chosen by AIC/BIC

Add a breakpoint at age=X (segmented regression) ONLY FOR OOCYTES, select X by AIC/BIC:
```{r}
age_break_try = 5:15

regression_break = function(age_break){
  age_break_scaled = (age_break-mean(mut$age_years))/sd(mut$age_years)
  mut$age_break_var = ( mut$age_years_scaled > age_break_scaled ) * ( mut$age_years_scaled - age_break_scaled ) * ( mut$tissue == "oocyte" )
  glmer_results_break <- glmer(mut_freq ~ age_years_scaled*tissue + age_break_var + (1|macaque_ID), 
                               weights = nt_sequenced, family = binomial(link = "logit"), 
                               control=glmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)), 
                               data = mut)
  return(c(age_break, summary(glmer_results_break)$AICtab[1:2]))
}

break_results = Reduce(rbind, lapply(age_break_try, regression_break))
colnames(break_results) = c('age', 'AIC', 'BIC')
row.names(break_results) = NULL
break_results
```

Best models:
```{r}
best_age = age_break_try[apply(break_results[,2:3], 2, which.min)]
names(best_age) = c('AIC', 'BIC')
best_age
```

Best model based on AIC and BIC has a break at years=9. Complete model:
$$E(Y)=\beta_0+\beta_1 age\_scaled+\beta_2H+\beta_3L+\beta_4O+\beta_5 I_{age>9} \times (age-9)\_scaled \times O+\beta_6 age\_scaled \times H+\beta_7 age\_scaled \times L+\beta_8 age\_scaled \times O$$
```{r}
age_break = 9
age_break_scaled = (age_break-mean(mut$age_years))/sd(mut$age_years)
mut$age_break_var = ( mut$age_years_scaled > age_break_scaled ) * ( mut$age_years_scaled - age_break_scaled ) * ( mut$tissue == "oocyte" )

glmer_results_break <- glmer(mut_freq ~ age_years_scaled*tissue + age_break_var + (1|macaque_ID), 
                             weights = nt_sequenced, family = binomial(link = "logit"), 
                             control=glmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)), 
                             data = mut)
summary(glmer_results_break)
```

Coefficients transforming back `age_years` to the original scale:
```{r}
beta_model = summary(glmer_results_break)$coefficients[,1]
mu = mean(mut$age_years)
sigma = sd(mut$age_years)
beta_break_rescaled_back = beta_model
beta_break_rescaled_back[c(2,6,7:9)] = beta_model[c(2,6,7:9)]/sigma
beta_break_rescaled_back[c(1,3:5)] = beta_model[c(1,3:5)] - beta_model[c(2,7:9)]*mu/sigma

as.matrix(beta_break_rescaled_back)
```

Odds ratio:
```{r}
odds_ratio_break = exp(beta_break_rescaled_back)
odds_ratio_break[1] = NA
as.matrix(odds_ratio_break)
```

Check residuals
```{r}
plot(glmer_results_break, col=c("green", "red", "blue", "black")[mut$tissue])
```

Compute pseudo_R2
```{r}
r.squaredGLMM(glmer_results_break, null = glmer_null)
r2_full_break <- r.squaredGLMM(glmer_results_break, null = glmer_null)[1,1]
r2_full_break
```

## Test difference with initial model
Test age_break_var
```{r}
anova(glmer_results_break, glmer_results)
anova(glmer_results_break, glmer_results)$`Pr(>Chisq)`
```

## Test random effect
Test random effect macaque_ID
```{r}
glmer_break_red <- glm(mut_freq ~ age_years_scaled*tissue + age_break_var, 
                       weights = nt_sequenced, family = binomial(link = "logit"), 
                       data = mut)
anova(glmer_results_break, glmer_break_red)
anova(glmer_results_break, glmer_break_red)$`Pr(>Chisq)`
```

Conditional modes of the random effects:
```{r fig.height=5, fig.width=6}
ranef_macaque_ID = ranef(glmer_results_break)$macaque_ID[,1]
summary(ranef_macaque_ID)

layout(matrix(c(1,1,2), nrow = 3))
hist(ranef_macaque_ID, main = 'Random effect macaque_ID', xlab = 'Conditional mode', xlim = c(-0.4, 0.4))
boxplot(ranef_macaque_ID, main = 'Random effect macaque_ID', xlab = 'Conditional mode', horizontal = TRUE, ylim = c(-0.4, 0.4))
```

Test if a random effect primate_center would also be important
```{r}
glmer_break_center <- glmer(mut_freq ~ age_years_scaled*tissue + age_break_var + (1|macaque_ID) + (1|primate_center), 
                            weights = nt_sequenced, family = binomial(link = "logit"), 
                            control=glmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)), 
                            data = mut)
anova(glmer_break_center, glmer_results_break)
```
It's not important at all (once macaque_ID is included)

## Test fixed effects
Test effect of age_years
```{r}
glmer_break_red <- glmer(mut_freq ~ tissue + (1|macaque_ID),
                         weights = nt_sequenced, family = binomial(link = "logit"), 
                         control=glmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)),
                         data = mut)
a=anova(glmer_results_break, glmer_break_red)
a$`Pr(>Chisq)`
```
Partial R2
```{r}
r2 <- r.squaredGLMM(glmer_break_red, null=glmer_null)[1,1]
(r2_full - r2) / (1-r2) * 100
```

Test effect of tissue
```{r}
glmer_break_red <- glmer(mut_freq ~ age_years_scaled + (1|macaque_ID),
                         weights = nt_sequenced, family = binomial(link = "logit"),
                         control=glmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 10000)),
                         data = mut)
a=anova(glmer_results_break, glmer_break_red)
a$`Pr(>Chisq)`
```
Partial R2
```{r}
r2 <- r.squaredGLMM(glmer_break_red, null=glmer_null)[1,1]
(r2_full - r2) / (1-r2) * 100
```

## Test slopes for different tissues
```{r}
coeffs_glmer = coeffs(glmer_results_break)
coeffs_glmer
```

Test muscle slope:
```{r}
beta1 = coeffs_glmer[2]
beta1/sigma

K = matrix(0, nrow = 1, ncol = length(coeffs_glmer))
K[1,c(2)] = 1/sigma
colnames(K) = names(coeffs_glmer)
rownames(K) = "slope_muscle"
summary(glht(glmer_results_break, linfct = K))
2*pnorm(-summary(glht(glmer_results_break, linfct = K))$test$tstat)
```

Test heart slope:
```{r}
beta1_6 = coeffs_glmer[2]+coeffs_glmer[7]
beta1_6/sigma

K = matrix(0, nrow = 1, ncol = length(coeffs_glmer))
K[1,c(2,7)] = 1/sigma
colnames(K) = names(coeffs_glmer)
rownames(K) = "slope_heart"
summary(glht(glmer_results_break, linfct = K))
2*pnorm(-summary(glht(glmer_results_break, linfct = K))$test$tstat)
```

Test liver slope:
```{r}
beta1_7 = coeffs_glmer[2]+coeffs_glmer[8]
beta1_7/sigma

K = matrix(0, nrow = 1, ncol = length(coeffs_glmer))
K[1,c(2,8)] = 1/sigma
colnames(K) = names(coeffs_glmer)
rownames(K) = "slope_liver"
summary(glht(glmer_results_break, linfct = K))
2*pnorm(-summary(glht(glmer_results_break, linfct = K))$test$tstat)
```

Model for oocytes ($H=0$, $L=0$ and $O=1$):
$$E(Y)=\left[\beta_0+\beta_4-(\beta_1+\beta_8)\frac{\mu_{age}}{\sigma_{age}}-9\frac{\beta_5I_{age>9}}{\sigma_{age}}\right]+\left(\frac{\beta_1+\beta_8+\beta_5 I_{age>9}}{\sigma_{age}}\right) age $$

Test if the effect of age is significant for $age \leq 9$:
$$H_0: \frac{\beta_1+\beta_8}{\sigma_{age}}=0 \qquad vs \qquad H_1: \frac{\beta_1+\beta_8}{\sigma_{age}} \neq 0$$
```{r}
beta1_8 = coeffs_glmer[2]+coeffs_glmer[9]
beta1_8/sigma

K = matrix(0, nrow = 1, ncol = length(coeffs_glmer))
K[1,c(2,9)] = 1/sigma
colnames(K) = names(coeffs_glmer)
rownames(K) = "slope_oocyte_less9"
summary(glht(glmer_results_break, linfct = K))
2*pnorm(-summary(glht(glmer_results_break, linfct = K))$test$tstat)
```

Test if the effect of age is significant for $age>9$:
$$H_0: \frac{\beta_1+\beta_8+\beta_5 I_{age>9}}{\sigma_{age}}=0 \qquad vs \qquad H_1: \frac{\beta_1+\beta_8+\beta_5 I_{age>9}}{\sigma_{age}} \neq 0$$
```{r}
beta1_8_5 = coeffs_glmer[2]+coeffs_glmer[9]+coeffs_glmer[6]
beta1_8_5/sigma

K = matrix(0, nrow = 1, ncol = length(coeffs_glmer))
K[1,c(2,9,6)] = 1/sigma
colnames(K) = names(coeffs_glmer)
rownames(K) = "slope_oocyte_more9"
summary(glht(glmer_results_break, linfct = K))
2*pnorm(-summary(glht(glmer_results_break, linfct = K))$test$tstat)
```


## Plot model prediction

```{r fig.height=7, fig.width=11}
options(repr.plot.width=11, repr.plot.height=7)

age_plot = seq(min(mut$age_years),max(mut$age_years),length.out = 50)
data_plot = data.frame(tissue = rep(levels(mut$tissue), each=50),
                       age_years = rep(age_plot, 4),
                       age_years_scaled = rep((age_plot-mean(mut$age_years))/sd(mut$age_years), 4))
data_plot$age_break_var = ( data_plot$age_years_scaled > age_break_scaled ) * ( data_plot$age_years_scaled - age_break_scaled ) * ( data_plot$tissue == "oocyte" )
data_plot$mut_freq = predict(glmer_results_break, data_plot, re.form=~0, type='response')

p <- ggplot(mut, aes(x=age_years, y=mut_freq, color=tissue)) + 
  geom_point(shape=16, size=4, alpha=0.9) +
  scale_color_manual(values=c("green", "red","blue", "black")) +
  theme_classic() +
  labs(title="",x="Age [years]",y="Mutation frequency [substitutions/bp]") +
  theme(axis.text.x = element_text(size=20, color="black"),axis.text.y = element_text(size=20, color="black"), axis.title.x = element_text(size=20), axis.title.y = element_text(size=20)) +
  guides(color=guide_legend(title="Tissue type")) +
  theme(legend.text=element_text(size=20)) +
  theme(legend.title=element_text(size=20, face="bold")) +
  theme(plot.title = element_text(size=20, lineheight=.8, face="bold", hjust = 0.5)) +
  
  geom_point(mut, mapping = aes(x = age_years, y = fitted(glmer_results), color=tissue), shape = 3, size = 4) + 
  geom_line(data_plot, mapping = aes(x = age_years, y = mut_freq, color=tissue), lwd = 1.5)

ggsave(filename="2021-04-03_mut_freq_d100_model_pred_breakAIC_oocyte.pdf", plot=p, , width = 11, height = 7)

p
```


```{r}
age_plot = seq(min(mut$age_years),max(mut$age_years),length.out = 50)
data_plot = data.frame(tissue = rep(levels(mut$tissue), each=50),
                       age_years = rep(age_plot, 4),
                       age_years_scaled = rep((age_plot-mean(mut$age_years))/sd(mut$age_years), 4))
data_plot$age_break_var = ( data_plot$age_years_scaled > age_break_scaled ) * ( data_plot$age_years_scaled - age_break_scaled ) * ( data_plot$tissue == "oocyte" )
data_plot$mut_freq = predict(glmer_results_break, data_plot, re.form=~0, type='response')

# bootstrap for computing CI using quantile approach
predFun <- function(fit){
  predict(fit, data_plot, re.form=~0, type='response')
}
merBoot = bootMer(glmer_results_break, FUN=predFun, nsim=100)
CIlevel = 0.95
data_plot$mut_freq_CIlower = apply(merBoot$t, 2, function(x) as.numeric(quantile(x, probs=(1-CIlevel)/2, na.rm=TRUE)))
data_plot$mut_freq_CIupper = apply(merBoot$t, 2, function(x) as.numeric(quantile(x, probs=(1+CIlevel)/2, na.rm=TRUE)))
```

```{r fig.height=7, fig.width=11}
options(repr.plot.width=11, repr.plot.height=7)

p <- ggplot(mut, aes(x=age_years, y=mut_freq, color=tissue)) + 
  geom_point(shape=16, size=4, alpha=0.9) +
  scale_color_manual(values=c("green", "red","blue", "black")) +
  theme_classic() +
  labs(title="",x="Age [years]",y="Mutation frequency [substitutions/bp]") +
  theme(axis.text.x = element_text(size=20, color="black"),axis.text.y = element_text(size=20, color="black"), axis.title.x = element_text(size=20), axis.title.y = element_text(size=20)) +
  guides(color=guide_legend(title="Tissue type")) +
  theme(legend.text=element_text(size=20)) +
  theme(legend.title=element_text(size=20, face="bold")) +
  theme(plot.title = element_text(size=20, lineheight=.8, face="bold", hjust = 0.5)) +
  
  geom_point(mut, mapping = aes(x = age_years, y = fitted(glmer_results), color=tissue), shape = 3, size = 4) + 
  geom_line(data_plot, mapping = aes(x = age_years, y = mut_freq, color=tissue), lwd = 1.5) +
  
  geom_ribbon(data_plot, mapping = aes(x=age_years, ymax=mut_freq_CIlower, ymin=mut_freq_CIupper), fill="gray", alpha=.25)

ggsave(filename="2021-12-23_mut_freq_d100_model_pred_breakAIC_oocyte_CI.pdf", plot=p, , width = 11, height = 7)

p
```

```{r fig.height=7, fig.width=11}
options(repr.plot.width=11, repr.plot.height=7)

age_plot = seq(min(mut$age_years),max(mut$age_years),length.out = 50)
data_plot = data.frame(tissue = rep(levels(mut$tissue), each=50),
                       age_years = rep(age_plot, 4),
                       age_years_scaled = rep((age_plot-mean(mut$age_years))/sd(mut$age_years), 4))
data_plot$age_break_var = ( data_plot$age_years_scaled > age_break_scaled ) * ( data_plot$age_years_scaled - age_break_scaled ) * ( data_plot$tissue == "oocyte" )
data_plot$mut_freq = predict(glmer_results_break, data_plot, re.form=~0, type='response')

p <- ggplot(mut, aes(x=age_years, y=mut_freq, color=tissue)) + 
  geom_point(shape=16, size=4, alpha=0.9) +
  scale_color_manual(values=c("green", "red","blue", "black")) +
  theme_classic() +
  labs(title="",x="Age [years]",y="Mutation frequency [substitutions/bp]") +
  theme(axis.text.x = element_text(size=20, color="black"),axis.text.y = element_text(size=20, color="black"), axis.title.x = element_text(size=20), axis.title.y = element_text(size=20)) +
  guides(color=guide_legend(title="Tissue type")) +
  theme(legend.text=element_text(size=20)) +
  theme(legend.title=element_text(size=20, face="bold")) +
  theme(plot.title = element_text(size=20, lineheight=.8, face="bold", hjust = 0.5)) +
  
  #geom_point(mut, mapping = aes(x = age_years, y = fitted(glmer_results), color=tissue), shape = 3, size = 4) + 
  geom_line(data_plot, mapping = aes(x = age_years, y = mut_freq, color=tissue), lwd = 1.5)

ggsave(filename="2021-04-03_mut_freq_d100_model_pred_breakAIC_oocyte_no_plus.pdf", plot=p, , width = 11, height = 7)

p
```


